Type theory in type theory with single substitutions

### 2024-09-18 types_2024_postproc/p.tex
stuff

The single substitution semantics is in some sense
too minimalistic: for example, consider a term $b$ which depends on a
context $\Gamma$ and an extra variable of type $A$, that is, $b :
\Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so that we add an extra $A$
variable \emph{before} the last one obtaining $ b[\p^+] :
\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p]) $ where $\p$ is weakening at
the end of the context, and $\p^+$ is weakening just before the last
variable in the context. Then, we instantiate the last $A[\p]$
variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$
denotes the last variable in the context (a.k.a\ De Bruijn index $0$).
We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary single substitution
model, but this equation holds in the single substitution syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In fact, we show that
the single substitution syntax of (\ref{eq:tt}) is equivalent to the
parallel substitution syntax of (\ref{eq:tt}), and the techniques that
we use are not specific to the SOGAT (\ref{eq:tt}), we expect
that they work for any SOGAT. Equivalence of the syntaxes means that
the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
general SOGAT, there are more models of the correponding single
substitution GAT than the parallel substitution GAT. Every parallel
model is also a single model, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
substitution models of (\ref{eq:tytm}) than parallel substitution
models. The situation is similar to that of extensional combinatory
calculus and lambda calculus, where the syntaxes are equivalent, and
every lambda model is a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship of monoids over a set $X$ and nil/cons algebras over $X$,
where every monoid is a nil/cons-algebra, but not the other way; it is
well-known that free monoids over $X$ (syntax for monoid over $X$) are
the same as lists of $X$ (the syntax for nil/cons algebras over $X$).

When showing that the parallel and single substitution syntaxes are
equivalent, we define the parallel operations by induction on the
single substitution syntax. One of the operations in the single
substitution syntax is the instantiation operation: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on types/terms, we have to provide
a method for these operations, but this can be difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal types and terms. Just as
$\beta$-normal forms don't distinguish between $\beta$-equal terms,
$\alpha$-normal forms don't include explicit substitutions.
$\alpha$-normal forms are not usual normal
forms in the sense that they are still quotiented by the
computation/uniqueness rules like $\beta$/$\eta$ for function
space. To derive the new induction principle for the syntax, we first
prove $\alpha$-normalisation.

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In its single substitution calculus, this can be
used to adapt some equations and remove some others, obtaining a more
minimalistic, but equivalent GAT.

>>>>

The single substitution semantics is in some sense too minimalistic:
for example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
\emph{derivable} in any model, but it is \emph{admissible} in the
syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ meaning that it is a term
which depends on a context $\Gamma$ and an extra variable of type $A$;
$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version
where we weakened \emph{before} the last variable; in
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
the last variable for the previous one.)  This is analogous to
e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
which do not hold in an arbitrary model, but they hold in the
syntax. In this paper, we show that for the particular SOGAT
(\ref{eq:tt}), the \emph{syntaxes} of the single and parallel
semantics are isomorphic. The techniques that we use for this proof
are not specific to the SOGAT (\ref{eq:tt}), we expect
that they work for any SOGAT. Isomorphism of the syntaxes means that
the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
general SOGAT, there are more models of the correponding single
substitution GAT than the parallel substitution GAT. Every parallel
model is also a single model, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
substitution models of (\ref{eq:tytm}) than parallel substitution
models (CwFs). The situation is similar to that of extensional combinatory
calculus and lambda calculus, where the syntaxes are equivalent, and
every lambda model is a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship of monoids over a set $X$ and nil/cons algebras over $X$,
where every monoid is a nil/cons-algebra, but not the other way; it is
well-known that free monoids over $X$ (syntax of monoid over $X$) are
the same as lists of $X$ (the syntax of nil/cons algebras over $X$).

When showing that the parallel and single substitution syntaxes are
equivalent, we define the parallel operations by induction on the
single substitution syntax. One of the operations in the single
substitution syntax is the instantiation operation: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on types/terms, we have to provide
a method for these operations, but this is difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal types and terms. Just as
$\beta$-normal forms don't distinguish between $\beta$-equal terms,
$\alpha$-normal forms don't include explicit substitutions.
$\alpha$-normal forms are not usual normal
forms in the sense that they are still quotiented by the
computation/uniqueness rules like $\beta$/$\eta$ for function
space. To derive the new induction principle for the syntax, we first
prove $\alpha$-normalisation.

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In its single substitution calculus, we use this to adapt some
equations and remove some others, obtaining a more minimalistic, but
isomorphic GAT.
